---
title: "Bivariate Moran's I — Spatial association between NegES and PosES"
author: "HT"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

Reproducible workflow in R using **spdep** to compute Bivariate Moran's I for grid-cell data.

**Usage:**

- Add your data files to a subfolder named `dataInput`
- Only edit code chunks marked **EDIT**

**Input data requirements:**
It can be raster or csv or shapefile. If input data is Shapefile with point geometry, you should have values to compare in a single shapefile with both results of to compare (e.g., pos and neg ES).
- `grid.shp` — geometry
- `grid.dbf` — attribute table with `NegES` and `PosES` columns
- `grid.shx` — shape index
- `grid.prj` — coordinate reference system (CRS)
- No missing values (or handle them explicitly) (zeros are accepted)

**Analytical logic:**

The analysis follows three sequential steps of increasing spatial specificity:

1. **Co-occurrence** (OR condition to cells selection) — do NegES and PosES share the same space?
2. **Pearson correlation** (AND condition to cells selection) — within shared space, are values correlated?
3. **Bivariate Moran's I** (AND condition to cells selection) — is that correlation spatially structured?


**Note: To make this analysis more meaningful we subset to presence-only cells before running the biv I test, to only look at variation among cells where values occur, not the presence/absence pattern itself (as we expect furthermore a lot of the studied area to be "empty"); otherwisewe would largely just confirm that ES don't occur everywhere in the Mediterranean, which we already knew!**

---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(terra)   # modern raster package
library(sf)
library(spdep)
library(here)
library(ggplot2)

```

---

## 1) Input data — EDIT THIS

```{r input EDIT}
fileName   <- here("dataInput", "CIMPAL-Meditteranean Sea.shp")  # edit file name
input_type <- "shp"    # "shp", "raster", or "csv"
input_crs  <- 4326     # only used for csv input
target_crs <- 3035     # reproject to this (must be metres); 3035 = ETRS89 LAEA for Mediterranean
```

---

## 2) Load & reproject

```{r load data}
grid <- switch(input_type,
  "raster" = as.points(rast(fileName)) |> st_as_sf(),
  "csv"    = st_as_sf(read.csv(fileName), coords = c("x", "y"), crs = input_crs),
  "shp"    = st_read(fileName),
  stop("input_type must be 'shp', 'raster', or 'csv'")
)

grid <- st_transform(grid, crs = target_crs)
cat("CRS units:", st_crs(grid)$units, "\n")
cat("Rows loaded:", nrow(grid), "\n")
```

---

## 3) Choose neighbour method — EDIT THIS

| `nb_method` | When to use |
|---|---|
| `"manual_dist"` | Regular grid, CRS in metres, you know the cell spacing |
| `"auto_dist"` | Regular grid but you don't want to hard-code the distance |
| `"knn"` | Best for regular grids — guarantees exactly k neighbours, no isolates |

```{r nb method EDIT}

nb_method <- "manual_dist" # edit, choose one from: "manual_dist" | "auto_dist" | "knn"

# Only used if nb_method = "manual_dist"
# e.g., 1000m = strictly adjacent (N/S/E/W); 1500m = includes diagonals (~1.41 km) on a 1 km grid
manual_threshold <- 15000 # edit

# Only used if nb_method = "knn"
# k = 4 → N/S/E/W only; k = 8 → N/S/E/W + 4 diagonals
knn_k <- 4 # edit
```

---

## 4) Co-occurrence summary (OR condition — descriptive)

Describes how NegES and PosES overlap spatially across all cells.

```{r co-occurrence summary}
grid_clean <- grid[!duplicated(st_geometry(grid)), ]
grid_or    <- grid_clean[grid_clean$NegES != 0 | grid_clean$PosES != 0, ]

n_both    <- nrow(grid_or[grid_or$NegES != 0 & grid_or$PosES != 0, ])
n_neg     <- nrow(grid_or[grid_or$NegES != 0 & grid_or$PosES == 0, ])
n_pos     <- nrow(grid_or[grid_or$NegES == 0 & grid_or$PosES != 0, ])
n_total   <- nrow(grid_or)

cat("=== Co-occurrence summary ===\n")
cat("Total cells with any ES signal:      ", n_total, "\n")
cat("Cells with NegES only:               ", n_neg, 
    "(", round(n_neg / n_total * 100, 1), "%)\n")
cat("Cells with PosES only:               ", n_pos, 
    "(", round(n_pos / n_total * 100, 1), "%)\n")
cat("Cells with BOTH (co-occurrence):     ", n_both, 
    "(", round(n_both / n_total * 100, 1), "%)\n")
```

---

## 5) Pearson correlation (AND condition)

Non-spatial correlation between NegES and PosES within co-occurrence cells.
This is a prerequisite check before spatial analysis — a high Bivariate Moran's I is only meaningful if the underlying non-spatial correlation is also strong.

```{r pearson}
grid_and <- grid_clean[grid_clean$NegES != 0 & grid_clean$PosES != 0, ]

pearson_r <- cor.test(grid_and$NegES, grid_and$PosES)

cat("=== Pearson correlation (co-occurrence cells only) ===\n")
cat("n cells:    ", nrow(grid_and), "\n")
cat("NegES — mean:", round(mean(grid_and$NegES), 3), 
              " sd:", round(sd(grid_and$NegES), 3), "\n")
cat("PosES — mean:", round(mean(grid_and$PosES), 3), 
              " sd:", round(sd(grid_and$PosES), 3), "\n")
cat("Pearson r:  ", round(pearson_r$estimate, 3), "\n")
cat("p-value:    ", pearson_r$p.value, "\n")
cat("95% CI:     ", round(pearson_r$conf.int[1], 3), 
              "—", round(pearson_r$conf.int[2], 3), "\n")

# Distribution plots
par(mfrow = c(1, 2))
hist(grid_and$NegES, main = "NegES distribution", xlab = "NegES", breaks = 50, col = "#ea4b36")
hist(grid_and$PosES, main = "PosES distribution", xlab = "PosES", breaks = 50, col = "#369b53")
par(mfrow = c(1, 1))
```

---

## 6) Build spatial weights (AND condition)

```{r build neighbours}
# ---- Helper: build neighbour object from chosen method ----
build_nb <- function(coords, method, manual_threshold, knn_k) {
  coords <- as.matrix(coords[, 1:2])
  storage.mode(coords) <- "double"

  if (method == "auto_dist") {
    k1       <- knn2nb(knearneigh(coords, k = 1))
    max_dist <- max(unlist(nbdists(k1, coords)))
    message("Auto threshold: ", round(max_dist), " m")
    dnearneigh(coords, d1 = 0, d2 = max_dist)

  } else if (method == "manual_dist") {
    dnearneigh(coords, d1 = 0, d2 = manual_threshold)

  } else if (method == "knn") {
    knn2nb(knearneigh(coords, k = knn_k))

  } else {
    stop("Unknown nb_method. Choose: 'auto_dist', 'manual_dist', or 'knn'.")
  }
}

# ---- Subset to co-occurrence cells ----
grid_presence <- grid_clean[grid_clean$NegES != 0 & grid_clean$PosES != 0, ]
cat("Co-occurrence cells:", nrow(grid_presence), "\n")

# ---- Extract centroids ----
coords_presence <- suppressWarnings(
  st_coordinates(st_centroid(grid_presence))[, 1:2]
)

# ---- Build initial neighbour structure ----
nb_presence <- build_nb(coords_presence, nb_method, manual_threshold, knn_k)
lw_presence <- nb2listw(nb_presence, style = "W", zero.policy = TRUE)

cat("CRS units:", st_crs(grid_presence)$units, "\n")
cat("Isolated cells before pruning:", sum(card(nb_presence) == 0), "\n")
```

```{r remove isolated cells}
# ---- Iteratively remove isolated cells ----
# Only relevant for distance-band methods; knn never produces isolates
repeat {
  isolated <- which(card(nb_presence) == 0)
  if (length(isolated) == 0) break

  if (nrow(grid_presence) - length(isolated) < 3) {
    warning("Too few presence cells remain after removing isolates — stopping early.")
    grid_presence <- grid_presence[-isolated, ]
    break
  }

  grid_presence   <- grid_presence[-isolated, ]
  coords_presence <- suppressWarnings(
    st_coordinates(st_centroid(grid_presence))[, 1:2]
  )
  nb_presence <- build_nb(coords_presence, nb_method, manual_threshold, knn_k)
  lw_presence <- nb2listw(nb_presence, style = "W", zero.policy = TRUE)
}

# ---- Verify ----
stopifnot(sum(card(nb_presence) == 0) == 0)

# ---- Extract variables ----
x <- grid_presence$NegES
y <- grid_presence$PosES

stopifnot(length(x) == length(y), 
          length(x) == nrow(grid_presence),
          length(lw_presence$neighbours) == nrow(grid_presence))

cat("Final cells for analysis:", nrow(grid_presence), "\n")
cat("Isolated cells remaining:", sum(card(nb_presence) == 0), "\n")
```

---

## 7) Global Bivariate Moran's I

```{r global morans i}
spdep::set.ZeroPolicyOption(TRUE)
res <- moran_bv(x, y, lw_presence, nsim = 999)

cat("=== Global Bivariate Moran's I ===\n")
cat("Bivariate Moran's I:", round(res$t0, 4), "\n")
cat("p-value:            ", (sum(res$t >= res$t0) + 1) / (length(res$t) + 1), "\n")
cat("Bias:               ", round(mean(res$t) - res$t0, 6), "\n")
cat("Std. error:         ", round(sd(res$t), 6), "\n")
```

**Output interpretation:**

| I value | Meaning |
|---|---|
| I > 0 | Similar values cluster together spatially |
| I < 0 | Opposing values cluster spatially |
| I ≈ 0 | No spatial relationship |

Permutation p-value `< 0.05` → significant spatial association.

---

## 8) Local Bivariate Moran's I (LISA)

Identifies *where* the spatial association is strongest.

| Quadrant | Ecological meaning |
|---|---|
| HH (red) | High NegES surrounded by high PosES — co-occurrence hotspots |
| LL (blue) | Low NegES surrounded by low PosES — marginal areas |
| HL (orange) | High NegES, low PosES neighbours — negative ES hotspots: Degradation without benefit — negative trade-off |
| LH (light blue) | Low NegES, high PosES neighbours — positive ES hotspots: Benefit without degradation — positive trade-off |

HL & LH are spatial trade-off zones.

```{r local LISA}
# ---- Spatially lag y ----
y_lag     <- lag.listw(lw_presence, y)
x_std     <- scale(x)[, 1]
y_lag_std <- scale(y_lag)[, 1]

# ---- Local statistic ----
local_bv              <- x_std * y_lag_std
grid_presence$local_bv <- local_bv

# ---- Significance via permutation ----
set.seed(123)
nsim <- 999

local_bv_sim <- replicate(nsim, {
  y_perm      <- sample(y)
  y_lag_p     <- lag.listw(lw_presence, y_perm)
  y_lag_std_p <- scale(y_lag_p)[, 1]
  x_std * y_lag_std_p
})

p_local                <- rowMeans(abs(local_bv_sim) >= abs(local_bv))
grid_presence$p_local  <- p_local

# ---- Classify LISA quadrants ----
grid_presence$lisa_class <- NA_character_
sig <- p_local < 0.05

grid_presence$lisa_class[sig & x_std >  0 & y_lag_std >  0] <- "HH"
grid_presence$lisa_class[sig & x_std <  0 & y_lag_std <  0] <- "LL"
grid_presence$lisa_class[sig & x_std >  0 & y_lag_std <  0] <- "HL"
grid_presence$lisa_class[sig & x_std <  0 & y_lag_std >  0] <- "LH"

cat("=== LISA cluster counts ===\n")
print(table(grid_presence$lisa_class, useNA = "ifany"))
```

The proportion of HH vs HL vs LH cells will tell whether co-occurrence hotspots or trade-off zones dominate the significant cells.

```{r spatial coherence}
# ---- Spatial coherence of significant clusters ----
sig_cells <- grid_presence[!is.na(grid_presence$lisa_class), ]

nb_lisa <- knn2nb(knearneigh(
  st_coordinates(st_centroid(sig_cells)), k = 1))

cat("Significant cells:          ", nrow(sig_cells), "\n")
cat("Spatial sub-graphs (patches):", attr(nb_lisa, "ncomp")$nc, "\n")
cat("Mean patch size:            ", round(nrow(sig_cells) / attr(nb_lisa, "ncomp")$nc, 1), "cells\n")
```

```{r cluster counts per LISA class}
# ---- Number of spatial sub-graphs per LISA class ----

lisa_classes <- c("HH", "LL", "HL", "LH")

for (cls in lisa_classes) {
  
  cells <- grid_presence[!is.na(grid_presence$lisa_class) & 
                           grid_presence$lisa_class == cls, ]
  
  if (nrow(cells) == 0) {
    cat(cls, "— no significant cells\n")
    next
  }
  
  nb_cls   <- knn2nb(knearneigh(
    st_coordinates(st_centroid(cells)), k = 1))
  n_patches <- attr(nb_cls, "ncomp")$nc
  
  cat(cls, "—",
      "cells:", nrow(cells), "|",
      "patches:", n_patches, "|",
      "mean patch size:", round(nrow(cells) / n_patches, 1), "\n")
}
```


```{r map lisa}
# ---- Reclassify for plotting (non-significant shown as grey) ----
grid_presence$lisa_plot <- ifelse(is.na(grid_presence$lisa_class),
                                  "Not significant",
                                  grid_presence$lisa_class)

ggplot(grid_presence) +
  geom_sf(aes(colour = lisa_plot), size = 0.8) +
  scale_colour_manual(
    values = c(
      "HH"              = "#fdae61", 
      "LL"              = "#abd9e9",
      "HL"              = "#ea4b36", 
      "LH"              = "#369b53",
      "Not significant" = "grey80"
    ),
    name = "LISA cluster",
    labels = c(
      "HH"              = "High–High (co-occurrence hotspot)",
      "LL"              = "Low–Low (marginal area)",
      "HL"              = "High NegES / Low PosES (-ve trade-off)",
      "LH"              = "Low NegES / High PosES (+ve trade-off)",
      "Not significant" = "Not significant"
    )
  ) +
  labs(
    title    = "Local Bivariate Moran's I (LISA)",
    subtitle = "NegES ~ spatially lagged PosES | p < 0.05",
    caption  = paste0(
      ifelse(nb_method == "manual_dist", 
             paste0(manual_threshold / 1000, " km distance band"), 
             paste0("knn k = ", knn_k)),
      " | EPSG:", target_crs)
  ) +
  theme_minimal()
```
Patches exploration:
```{r patches size}
# Size distribution of patches
comp <- n.comp.nb(nb_lisa)
patch_sizes <- sort(table(comp$comp.id), decreasing = TRUE)
cat("Largest patches (top 10):\n")
print(head(patch_sizes, 10))
cat("Singletons (patch size = 1):", sum(patch_sizes == 1), "\n")
cat("Patches of size >= 10:", sum(patch_sizes >= 10), "\n")
```
```{r save map}
# ---- Save LISA map for publication ----

# Recreate plot as object
lisa_map <- ggplot(grid_presence) +
  geom_sf(aes(colour = lisa_plot), size = 0.8) +
  scale_colour_manual(
    values = c(
      "HH"              = "#fdae61", 
      "LL"              = "#abd9e9",
      "HL"              = "#ea4b36", 
      "LH"              = "#369b53",
      "Not significant" = "grey80"
    ),
    name = "LISA cluster",
    labels = c(
      "HH"              = "High–High (co-occurrence hotspot)",
      "LL"              = "Low–Low (marginal area)",
      "HL"              = "High NegES / Low PosES (-ve trade-off)",
      "LH"              = "Low NegES / High PosES (+ve trade-off)",
      "Not significant" = "Not significant"
    )
  ) +
  labs(
    title    = "Local Bivariate Moran's I (LISA)",
    subtitle = "NegES ~ spatially lagged PosES | p < 0.05",
    caption  = paste0(
      ifelse(nb_method == "manual_dist",
             paste0(manual_threshold / 1000, " km distance band"),
             paste0("knn k = ", knn_k)),
      " | EPSG:", target_crs)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = "right",
    legend.title      = element_text(face = "bold", size = 10),
    legend.text       = element_text(size = 9),
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(size = 10, colour = "grey40"),
    plot.caption      = element_text(size = 8, colour = "grey50"),
    panel.grid.major  = element_line(colour = "grey90"),
    panel.grid.minor  = element_blank()
  )

# ---- Export ----
# TIFF — preferred by most journals (300 dpi, CMYK-compatible)
ggsave(
  filename = here("outputs", "LISA_map.tiff"),
  plot     = lisa_map,
  width    = 180,   # mm — standard single-column width; use 90 for half-page
  height   = 140,   # mm — adjust to match your study area aspect ratio
  units    = "mm",
  dpi      = 300,
  compression = "lzw"  # lossless compression for TIFF
)

# PDF — vector format, ideal for resizing without quality loss
ggsave(
  filename = here("outputs", "LISA_map.pdf"),
  plot     = lisa_map,
  width    = 180,
  height   = 140,
  units    = "mm"
)

cat("Maps saved to outputs/ folder:\n")
cat("  LISA_map.tiff — 300 dpi, LZW compressed (for journal submission)\n")
cat("  LISA_map.pdf  — vector format (for editing in Illustrator/Inkscape)\n")
```
```


---

### Common pitfalls

- ❌ Mismatched grids → invalid comparison
- ❌ Isolated cells with no neighbours → use `zero.policy = TRUE` and prune iteratively
- ❌ Strong spatial autocorrelation inflating I → always use permutations (`nsim = 999`)
- ❌ OR condition with many zeros → use AND condition for Bivariate Moran's I; OR condition for descriptive co-occurrence only
- ❌ Wrong CRS → reproject to a metric CRS (EPSG:3035 for Mediterranean) before building neighbours
- ❌ Interpreting I as a difference → it measures **association**, not difference
